\chapter{Related Work}
\label{related_work}

This chapter surveys empirical studies that have measured the energy consumption and carbon emissions of machine learning systems. Each section closes by identifying what prior studies leave unmeasured or unresolved, collectively motivating the specific design choices of this study.

% =============================================================
\section{Empirical Energy Benchmarks for Machine Learning}
\label{sec:related_benchmarks}

As empirical benchmarking has replaced theoretical complexity analysis as the primary mode of model evaluation, a body of literature has emerged that measures the actual energy consumption and carbon emissions of training machine learning systems. This literature has, however, grown unevenly: the studies that sparked the Green \gls{ai} movement focused almost exclusively on large-scale deep learning, while the empirical benchmarks covering tabular data and classical algorithms ignored energy entirely.

The case for treating energy as a first-class evaluation metric was made most forcefully by early measurements of \gls{dl} workloads. \citeauthor{strubellEnergyPolicyConsiderations2019a} \cite{strubellEnergyPolicyConsiderations2019a} were among the first to directly measure hardware power interfaces during \gls{nlp} model training, finding that a single large transformer architecture could emit as much \gls{co2eq} as five cars over their entire lifespans and helping to establish the Green \gls{ai} research agenda. Follow-up work examined the structural drivers of this footprint. \citeauthor{dodgeMeasuringCarbonIntensity2022} \cite{dodgeMeasuringCarbonIntensity2022} expanded the empirical scope to 16 cloud computing regions, demonstrating that hardware choice and grid carbon intensity are as consequential as algorithmic decisions, with the same training run emitting between 7{,}000 and 26{,}000 grams of CO\textsubscript{2} depending on the data center. \citeauthor{desislavovTrendsAIInference2023} \cite{desislavovTrendsAIInference2023} extended this perspective to the inference phase, showing that the operational footprint of deployed models can rival the cost of training. A step toward more modest workloads was taken by \citeauthor{hendersonSystematicReportingEnergy2020} \cite{hendersonSystematicReportingEnergy2020}, whose experiment-impact-tracker framework demonstrated that everyday reinforcement learning experiments on standard hardware generate clearly measurable emissions, establishing ecological efficiency as a concern for ordinary research rather than exclusively for large-scale industrial training. Despite this broadening scope, the entire strand of Green \gls{ai} benchmarking has remained focused on neural networks trained on unstructured data such as text and images.

The empirical literature on tabular data took a different path entirely. A series of large-scale benchmarks asked which model family generalizes best to structured, heterogeneous datasets and consistently found that classical algorithms hold their own against deep learning. \citeauthor{grinsztajnWhyTreebasedModels2022a} \cite{grinsztajnWhyTreebasedModels2022a} evaluated 45 tabular datasets and found tree-based models such as \gls{rf} and \gls{xgb} to outperform deep neural networks, a result echoed by \citeauthor{gorishniyRevisitingDeepLearning2021a} \cite{gorishniyRevisitingDeepLearning2021a} in a separate benchmark that reaffirmed the dominance of gradient boosting on this data type. Both studies, however, treated energy as entirely outside their scope: runtime was noted as a secondary metric, but physical power consumption and carbon emissions went unmeasured.

Only recently have some studies begun to bridge these two strands. \citeauthor{rodriguezEvaluatingEnergyConsumption2024} \cite{rodriguezEvaluatingEnergyConsumption2024} evaluated software-based energy trackers against physical hardware meters, measuring classical \gls{ml} algorithms alongside \gls{dl} models as part of their methodology. \citeauthor{verdecchiaDataCentricGreenAI2022a} \cite{verdecchiaDataCentricGreenAI2022a} took a data-centric approach, empirically measuring how reducing data points and features affects the training energy of six \gls{ai} algorithms and finding that data-centric dataset modifications can substantially reduce training energy, though the accuracy cost varies depending on whether data points or features are reduced. These contributions are important, but they remain limited to datasets of a few thousand instances and do not examine how energy and accuracy co-evolve across the orders-of-magnitude variation in dataset size that characterizes real-world applications.

\paragraph{Research Gap.}
The existing literature therefore reveals a clear gap at the intersection of two well-developed but isolated fields. Where Green \gls{ai} benchmarks have established energy as a measurable, consequential quantity, they have done so exclusively for deep learning on unstructured data \parencite{strubellEnergyPolicyConsiderations2019a, dodgeMeasuringCarbonIntensity2022}. Where experimental benchmarks have rigorously evaluated classical algorithms on tabular data, they have focused entirely on predictive accuracy \parencite{grinsztajnWhyTreebasedModels2022a, gorishniyRevisitingDeepLearning2021a}. A systematic analysis of how energy consumption, carbon emissions and predictive accuracy jointly scale with dataset size for classical classification algorithms remains absent.

% =============================================================
\section{Dataset Size Effects on Performance and Energy Consumption}
\label{sec:related_dataset_size}

The tension between data volume and model quality has a specific shape. \citeauthor{schwartzGreenAI2020} \cite{schwartzGreenAI2020} conceptualize this in their Equation of Red \gls{ai}, which models the computational cost of a result as growing linearly with training dataset size. At the same time, accuracy does not grow linearly with data: empirical learning curves consistently follow an inverse power law, meaning that performance gains become increasingly marginal as the training set expands \cite{koshuteRecommendingTrainingSet2021}. The practical consequence is that energy costs scale proportionally with data volume while accuracy returns diminish, making large datasets progressively less efficient investments. \citeauthor{koshuteRecommendingTrainingSet2021} \cite{koshuteRecommendingTrainingSet2021} formalize the point where these curves cross as the Sufficient Training Set Size (STSS): the minimum volume beyond which additional data is ecologically unjustifiable. A complementary finding from \citeauthor{althnianImpactDatasetSize2021} \cite{althnianImpactDatasetSize2021}, based on medical classification tasks, adds that the key factor is not absolute size but representativeness, meaning that a small, well-sampled dataset can generalize as well as a much larger but biased one.

The energy implications of dataset reduction have been directly and concretely measured by \citeauthor{verdecchiaDataCentricGreenAI2022a} \cite{verdecchiaDataCentricGreenAI2022a}, who trained six \gls{ai} algorithms while systematically reducing both data points and features in 10\% increments. Two distinct patterns emerge from their results. Reducing the number of features yields energy savings of up to 76\% while largely preserving predictive performance, making it a near-cost-free optimization. Reducing the number of data points produces larger savings, up to 92.16\% for \gls{rf}, but at a meaningful accuracy cost that must be weighed explicitly. The effect is also algorithm-specific: \gls{knn} achieves a maximum reduction of only 61.72\% from fewer rows, while ensemble methods such as \gls{rf} are considerably more sensitive to training set size. Beyond dataset modifications, the study finds that switching to a less complex algorithm can alone yield energy savings of up to two orders of magnitude, establishing algorithm selection as one of the most powerful levers for ecological efficiency.

\paragraph{Research Gap.}
These findings rest, however, on a single dataset of 5{,}574 instances, run exclusively on a standard quad-core \gls{cpu} and without \gls{gpu}-accelerated or \gls{dl} baselines. Whether the same energy-accuracy patterns hold at the scale of millions of instances and whether they generalize across fundamentally different model families, remains empirically unexplored.

% =============================================================
\section{CPU vs.\ GPU Efficiency: Empirical Evidence}
\label{sec:related_cpu_gpu}

Whether \gls{gpu} training is more energy-efficient than \gls{cpu} training cannot be answered by measuring power draw alone. A \gls{gpu} that draws five times the wattage of a \gls{cpu} can still consume less total energy, if it finishes the same training task more than five times faster. The relevant quantity is energy, specifically power integrated over time and how that relationship plays out depends heavily on the algorithm and the dataset \cite{pattersonCarbonEmissionsLarge2021}.

The power side of this equation is well established empirically. \citeauthor{dodgeMeasuringCarbonIntensity2022} \cite{dodgeMeasuringCarbonIntensity2022} report observed \gls{gpu} draws of 250\,W to 350\,W under load from cloud training environments, compared to 10\,W to 150\,W for \glspl{cpu}. During \gls{bert} training, a workload representative of high-utilization deep learning, the \gls{gpu} alone accounted for 74\% of total system electricity. For \gls{nn} architectures of this type, the elevated power draw is more than offset by execution speed: the parallelism of deep learning maps efficiently onto \gls{gpu} hardware and training that would take days on a \gls{cpu} completes in hours. For this class of workload, the \gls{gpu} is the more energy-efficient device.

Classical \gls{ml} algorithms on tabular data present a different picture. \citeauthor{mitchellAcceleratingXGBoostAlgorithm2017} \cite{mitchellAcceleratingXGBoostAlgorithm2017} measured \gls{gpu} acceleration for \gls{xgb} specifically, finding speedups of 3$\times$ to 6$\times$ over a four-core desktop \gls{cpu} and 1.2$\times$ over a dual-socket server with 24 \gls{cpu} cores combined. These gains are real but considerably smaller than those achieved for \glspl{nn}. Given the \gls{gpu}'s substantially higher power draw, a 3$\times$ to 6$\times$ speedup may not be enough to recover the additional energy cost and Mitchell and Frank measured only execution time, not energy consumption.

\paragraph{Research Gap.}
The question of where the break-even point lies for algorithms like \gls{xgb} therefore remains open. A dataset may exist below which the \gls{cpu} is the ecologically preferable device and above which the \gls{gpu}'s speedup compensates for its power draw, but this has not been empirically established for gradient boosting on tabular data.

% =============================================================
\section{Measurement Tool Accuracy: Software Estimation vs.\ Hardware Sensors}
\label{sec:related_measurement}

As Green \gls{ai} research has grown, so has demand for accessible tools that quantify the energy consumption and carbon emissions of model training. The emerging ecosystem of software trackers and analytical estimators has simplified emissions measurement. However, recent empirical work reveals a significant problem: different tools often report strikingly different results for the same experiment. These discrepancies are not random but represent systematic errors in the underlying estimation models.

Several measurement approaches exist, ranging from hardware power meters at the wall outlet to on-chip sensors and software estimation models \cite{rodriguezEvaluatingEnergyConsumption2024}. In practice, cost and access constraints push most researchers toward estimation-based tools, despite evidence that these introduce substantial and directional errors.

How large these errors can be is documented by \citeauthor{bannourEvaluatingCarbonFootprint2021} \cite{bannourEvaluatingCarbonFootprint2021}, who compared six tracking tools on the same \gls{nlp} training runs. Tools relying on analytical models and user-supplied parameters systematically overestimated consumption, while those reading directly from hardware sensors produced more consistent results. The underlying reason for this divergence is that analytical approaches cannot capture the dynamic fluctuations of real hardware. \citeauthor{hendersonSystematicReportingEnergy2020} \cite{hendersonSystematicReportingEnergy2020} demonstrated this empirically: \gls{tdp}-based estimates significantly over- or underestimate actual energy compared to sensor readings, because they ignore real-time utilization shifts and dynamic memory effects and \glspl{flop}-based estimates often show little correlation with actual consumption across architectures due to hardware-specific firmware behaviors. \citeauthor{rodriguezEvaluatingEnergyConsumption2024} \cite{rodriguezEvaluatingEnergyConsumption2024} confirmed this pattern in practice, finding that purely analytical tools consistently deviated from measured ground truth, while sensor-based tools remained closer, though still subject to platform-specific constraints.

More recent work has moved toward direct validation of widely adopted trackers against physical ground truth. \citeauthor{fischerGroundTruthingAIEnergy2025} \cite{fischerGroundTruthingAIEnergy2025} compared CodeCarbon against external power meters across a range of workloads, finding that CodeCarbon consistently underestimates total system power draw by 20 to 30\%, as software profilers cannot account for cooling overhead or power supply inefficiencies. Critically, this underestimation is substantially worse for \gls{cpu}-only workloads than for \gls{gpu} workloads, with relative errors exceeding \gls{gpu}-equivalent errors by more than an order of magnitude in some configurations, a finding with direct consequences for any study benchmarking classical \gls{ml}, which runs predominantly on \gls{cpu}.

% =============================================================
\section{Seasonal and Geographic Carbon Intensity Effects}
\label{sec:related_carbon_intensity}

The carbon footprint of a machine learning experiment is not determined solely by algorithmic and hardware choices. Because electricity grids draw on different fuel mixes depending on geography, time of day and season, two identical training runs executed on different infrastructure can produce vastly different emissions, making geographic and temporal factors a distinct dimension of Green \gls{ai} research.

The geographic component of this variability has been measured at scale. \citeauthor{dodgeMeasuringCarbonIntensity2022} \cite{dodgeMeasuringCarbonIntensity2022} documented factors of up to 40 between the most and least carbon-intensive cloud regions, establishing workload placement as a practical lever for emissions reduction alongside hardware and algorithm choice.

Temporal variation tells a more nuanced story. \citeauthor{dodgeMeasuringCarbonIntensity2022} \cite{dodgeMeasuringCarbonIntensity2022} find that diurnal variation is meaningful: starting a training run at midnight rather than during peak solar generation hours can increase emissions by up to 8\% in certain regions and they propose time-shifting or dynamically pausing workloads based on real-time grid data. Seasonal variation, however, proves negligible across their 12-month tracking of globally distributed cloud instances, leading them to conclude that month-to-month emissions differences within a single cloud region are small.

\paragraph{Research Gap.}
This conclusion may not transfer to local electricity grids with a high share of seasonally dependent renewables, which can exhibit considerably stronger swings than global cloud averages suggest. 

% =============================================================
\section{Hyperparameter Tuning Efficiency}
\label{sec:tuning_efficiency}

To achieve state-of-the-art predictive performance, \gls{ml} models require extensive hyperparameter tuning. Since this process involves training architectures multiple times across various configurations, a critical tension emerges between the ecological cost of finding the optimal configuration and the performance benefits it yields.

The Red \gls{ai} paradigm has normalized evaluating only the final, best-performing model, a practice that obscures the computational budget spent during the development and tuning phases \cite{schwartzGreenAI2020}. Practical measurements reveal that this hidden cost can be substantial. \citeauthor{strubellEnergyPolicyConsiderations2019a} \cite{strubellEnergyPolicyConsiderations2019a}, in an empirical case study tracking the development of a \gls{nlp} pipeline, quantified this overhead directly. A single tuning cycle evaluating 24 configurations multiplied the computational cost of the baseline model by a factor of 24. Further the complete development process, which required training 4{,}789 model variants to find the optimal hyperparameters, consumed 239{,}942 \gls{gpu} hours in total. This demonstrates that the tuning phase can multiply the baseline energy footprint of a single model by several orders of magnitude.

To reduce these impacts, researchers must move away from exhaustive methods like grid search. \citeauthor{lacosteQuantifyingCarbonEmissions2019} \cite{lacosteQuantifyingCarbonEmissions2019} explicitly state that grid search is highly inefficient in terms of both model performance and environmental impact and that replacing it with random or model-based search strategies can accelerate convergence and directly reduce carbon emissions. Experimental benchmarks support this: \citeauthor{bergstraRandomSearchHyperParameter2012} \cite{bergstraRandomSearchHyperParameter2012} show that as few as 8 to 32 random trials can consistently match or outperform exhaustive grid search depending on the size of the search space. Going further, \citeauthor{bergstraAlgorithmsHyperParameterOptimization2011} \cite{bergstraAlgorithmsHyperParameterOptimization2011} illustrate the practical scale of this problem on a Deep Belief Network with 32 hyperparameters: evaluating only two values per parameter under grid search yields over four billion combinations, which at roughly one hour of training each would require nearly half a million years of compute. By distributing asynchronous \gls{tpe} proposals across five \glspl{gpu}, they evaluated only 200 configurations in 24 hours of wall time and identified a result that outperformed expert-tuned baselines \cite{bergstraAlgorithmsHyperParameterOptimization2011}.

\paragraph{Research Gap.}
While the literature clearly documents the empirical tuning costs for \gls{dl} models in \gls{nlp} and Computer Vision \cite{strubellEnergyPolicyConsiderations2019a}, there is a lack of equivalent benchmarks for classical \gls{dl} on tabular data. How much energy and CO\textsubscript{2} can be saved by reducing the number of tuning trials for algorithms like \gls{rf} or \gls{xgb} and at what trial count the marginal accuracy gain ceases to justify the ecological cost, remains unquantified.

% =============================================================
\section{Lifecycle Emissions: The Ecological Cost of Deployment and Inference}
\label{sec:related_inference}

While the \gls{ml} community has heavily scrutinized the energy consumption and carbon emissions of the training phase, this narrow focus risks obscuring the true ecological cost of an \gls{ai} system's complete lifecycle. A critical tension thus emerges: does the cumulative energy consumed during the deployment phase eventually outweigh the initial, highly-publicized training footprint?

Industry-scale empirical studies demonstrate that the inference phase is often the dominant factor in a model's total energy consumption. As models are integrated into everyday applications, they are queried millions of times, creating a multiplicative effect where the ongoing energy demand of continuous deployment quickly eclipses the one-time cost of training \cite{desislavovTrendsAIInference2023}. This dominance is reflected in industry estimates cited by \citeauthor{pattersonCarbonEmissionsLarge2021} \cite{pattersonCarbonEmissionsLarge2021}, who report that both NVIDIA and Amazon Web Services estimated that 80 to 90\% of total machine learning computing demand in large-scale deployed systems is attributable to inference. \citeauthor{wuSustainableAIEnvironmental2022} \cite{wuSustainableAIEnvironmental2022} provide further empirical support at the production level, finding that for deployed language models at Meta, the inference phase accounts for 65\% of total lifecycle resource consumption, significantly outweighing the training phase.

Consequently, recent research has begun explicitly measuring the inference footprint of pre-trained models at deployment scale. By systematically benchmarking energy consumption across diverse model architectures, \citeauthor{luccioniPowerHungryProcessing2024} \cite{luccioniPowerHungryProcessing2024} found that multipurpose generative models require orders of magnitude more energy per inference than task-specific systems, even when controlling for the number of model parameters. They also formalize the concept of a cost parity point: the number of predictions at which cumulative inference emissions surpass the initial training footprint, providing a concrete metric for lifecycle analysis. Accurately evaluating this break-even requires accounting for hardware specializations, as \citeauthor{desislavovTrendsAIInference2023} \cite{desislavovTrendsAIInference2023} note that inference efficiency depends heavily on whether a model can fully exploit the underlying accelerator's optimizations.

\paragraph{Research Gap.}
Despite the growing acknowledgment of inference costs, the current literature exhibits a significant blind spot. The lifecycle analyses and inference benchmarks provided by these studies rely almost exclusively on complex \gls{dl} architectures applied to unstructured data: \citeauthor{luccioniPowerHungryProcessing2024} \cite{luccioniPowerHungryProcessing2024} evaluate only text and image modalities, \citeauthor{desislavovTrendsAIInference2023} \cite{desislavovTrendsAIInference2023} focus on ImageNet and GLUE benchmarks, \citeauthor{pattersonCarbonEmissionsLarge2021} \cite{pattersonCarbonEmissionsLarge2021} restrict their analysis to large \gls{nlp} models and \citeauthor{wuSustainableAIEnvironmental2022} \cite{wuSustainableAIEnvironmental2022} examine \gls{dl} recommendation and language models. For classical \gls{dl} algorithms operating on tabular data, direct experimental evidence on lifecycle emissions is near-absent. It therefore remains unquantified at what number of predictions the cumulative inference CO\textsubscript{2} footprint of a model such as \gls{lr} or \gls{rf} surpasses its training footprint and how this lifecycle scaling compares across algorithms of varying complexity.

% =============================================================
\section{Composite Efficiency Metrics: Balancing Performance and Ecological Cost}
\label{sec:related_metrics}

The singular focus on predictive accuracy in machine learning has historically obscured the underlying computational costs of model training, raising a critical methodological question: How can researchers systematically evaluate both model quality and environmental footprint within a single, standardized framework?

To counteract the tendency to achieve marginal accuracy gains through massive computational expenditure, \citeauthor{schwartzGreenAI2020} \cite{schwartzGreenAI2020} argue that efficiency must be elevated to a primary evaluation criterion alongside accuracy. Measuring algorithmic progress exclusively through predictive performance is fundamentally incomplete. \citeauthor{hernandezMeasuringAlgorithmicEfficiency2020} \cite{hernandezMeasuringAlgorithmicEfficiency2020} demonstrate that true progress must be quantified by holding performance constant and measuring the reduction in computational cost required to achieve it. Taken together, these theoretical arguments advocate for composite metrics that mathematically integrate compute costs with model capabilities to provide a holistic view of algorithmic efficiency.

In practice, the machine learning community has already recognized the necessity of such combined measurements. \citeauthor{colemanAnalysisDAWNBenchTimetoAccuracy2019} \cite{colemanAnalysisDAWNBenchTimetoAccuracy2019} introduced composite metrics such as Time-to-Accuracy and Cost-to-Accuracy in the DAWNBench benchmark, demonstrating that the community accepts end-to-end metrics that evaluate the time and financial budget required to reach a predefined quality threshold. Building upon this precedent, the MLPerf benchmark suite \cite{mattsonMLPerfTrainingBenchmark2020} formally adopted Time-to-Train as its primary evaluation metric, deliberately shifting the focus away from raw accuracy towards the holistic efficiency of the training system.

While these benchmarks validate the use of composite metrics, their efficiency components rely exclusively on execution time or financial cost \parencite{colemanAnalysisDAWNBenchTimetoAccuracy2019, mattsonMLPerfTrainingBenchmark2020}. As established in previous sections, neither time nor cost necessarily correlates with the actual carbon footprint, particularly when hardware power draw and regional grid carbon intensities vary \cite{wrightEfficiencyNotEnough2025}. Recent theoretical frameworks have begun to recognize this shortcoming by proposing conceptual metrics, such as the Job Quality per Cost Rate (JQCR) \cite{eilamMethodologyFrameworkAI2023}, which aims to balance the quality of a computing job against its sustainability costs \cite{eilamMethodologyFrameworkAI2023}. However, this approach remains strictly conceptual and lacks a defined mathematical formula linking specific accuracy measurements to carbon emissions \cite{eilamMethodologyFrameworkAI2023}. 

\paragraph{Research Gap.}
To date, no established, mathematically operationalized composite metric explicitly incorporates a direct ecological component alongside predictive accuracy.