\chapter{Methodology}
\label{experimental_design}
This chapter outlines the design decisions underlying the empirical evaluation of this study. It describes the research questions, the selected datasets and models, the evaluation metrics, the validation strategy, and the hyperparameter tuning approach that together form the methodological foundation for the experiments conducted in \Cref{implementation}.

\section{Research Questions}
\label{sec:research_questions}
The main research question that this study aims to answer is: \textit{How does the trade-off between energy consumption and predictive performance vary across Logistic Regression, Random Forest, XGBoost and \gls{mlp} models when the complexity of the models and the size of the dataset are systematically varied?}
The following sub-questions will also be addressed in the course of this thesis:
\begin{itemize}
    \item Under what conditions do the resource consumption of more complex models exceed the achievable accuracy gain over simpler baselines?
    \item How does energy consumption and CO\textsubscript{2} output scale with increasing neural network complexity (number of layers and neurons)?
    \item In what way does energy consumption differ between \gls{cpu}-based and \gls{gpu}-based training for equivalent tasks?
    \item How does reducing the number of training instances on large datasets affect model accuracy and energy consumption across different algorithms?
    \item At what prediction volume does the cumulative inference CO\textsubscript{2} footprint of a model surpass its training footprint, and how does this break-even point differ across model families?
\end{itemize}


% =============================================================
\section{Datasets}
\label{experimental_dataset}
Dataset size is a primary driver of both computational demand and energy consumption during model training \cite{rodriguezEvaluatingEnergyConsumption2024}. To evaluate how this relationship varies across different scales, three classification datasets were selected that span several orders of magnitude in size. Classification was chosen as the task type since it is well-established across all selected models and allows for direct comparison of predictive performance via standardized metrics such as accuracy and weighted F1 score. The datasets are summarized in Table~\ref{tab:datasets}.

\begin{table}[htbp]
\centering
\caption{Overview of datasets used in this analysis}
\label{tab:datasets}
\begin{tabular}{lrrrl}
\toprule
\textbf{Dataset} & \textbf{Instances} & \textbf{Features} & \textbf{Classes} & \textbf{Scale} \\
\midrule
Wine \cite{stefanaeberhardWine1992} & 178 & 13 & 3 & Small \\
Default of Credit Card Clients \cite{i-chengyehDefaultCreditCard2009} & 30,000 & 23 & 2 & Medium \\
HIGGS \cite{baldiSearchingExoticParticles2014} & 11,000,000 & 28 & 2 & Large \\
\bottomrule
\end{tabular}
\end{table}

The Wine dataset contains 178 instances across 13 features, where each feature describes a chemical property of wines from the same region in Italy, derived from three different cultivars.

The Default of Credit Card Clients (hereafter: Credit) dataset contains 30,000 instances across 23 features and focuses on predicting payment default among credit card clients in Taiwan.

The HIGGS dataset is the largest dataset used in this study, comprising 11,000,000 instances across 28 features, seven of which are high-level features derived by physicists to help discriminate between the two classes \cite{whitesonHIGGS2014}. The classification goal is to distinguish between signal events produced by Higgs bosons and background noise processes.

These three datasets vary significantly in size, enabling a systematic evaluation of how model performance and energy consumption scale across different levels of data complexity. The jump from 30,000 to 11,000,000 instances is large, spanning nearly three orders of magnitude. This gap is addressed by the dataset scaling experiment described in \Cref{sec:experimental_protocol}, which trains models on seven intermediate HIGGS subset sizes ranging from 1,000 to 11,000,000 instances and thereby provides a continuous view of how performance and emissions evolve across the full size range.

\section{Models}
\label{design_models}
Selecting models of varying complexity is central to answering the research questions of this study. A model that achieves competitive accuracy with significantly lower energy consumption represents a more ecologically efficient choice, but this trade-off can only be quantified if the comparison spans a meaningful range of complexity. For this reason, four classification models were selected, ranging from a linear baseline to tree-based ensemble methods and a neural network, covering low, medium, and high computational demand respectively.

\gls{lr} serves as the lightweight baseline in this study. Despite its simplicity as a linear classifier, it often achieves competitive \gls{acc}, making it well suited to quantifying whether the additional resource consumption of more complex models yields a meaningful gain in predictive performance.

\gls{rf} was selected as a mid-complexity representative of tree-based ensemble methods. Relevant to this study, Random Forest does not support \gls{gpu}-accelerated training and is therefore run exclusively on the \gls{cpu}.

Building on the tree-based approach, \gls{xgb} was chosen as a second mid-complexity model, with the key distinction that it also supports \gls{gpu}-accelerated training. This makes it possible to directly compare the energy consumption and training time of \gls{cpu}- and \gls{gpu}-based tree ensemble methods under identical conditions, which is central to one of the research questions of this study. The suitability of \gls{gpu}-accelerated XGBoost for datasets at this scale has been demonstrated by Mitchell and Frank, who show that the histogram-based \gls{gpu} implementation can process a dataset of 10 million instances with 28 features entirely within \gls{gpu} memory \cite{mitchellAcceleratingXGBoostAlgorithm2017}.

The \gls{mlp} was selected as the most complex model, representing the neural network category. Its highly configurable architecture, particularly the number of hidden layers and neurons per layer, makes it well suited for the architectural complexity experiments conducted in this study, where these parameters are systematically varied to assess their impact on predictive performance and energy consumption.

\begin{table}[htbp]
\centering
\caption{Overview of classification models used in this study}
\label{tab:models}
\begin{tabular}{llll}
\toprule
\textbf{Model} & \textbf{Type} & \textbf{Device} & \textbf{Role} \\
\midrule
\gls{lr} & Linear & \gls{cpu} & Baseline \\
\gls{rf}  & Ensemble & \gls{cpu} & Mid-complexity \\
\gls{xgb} & Gradient Boosting & \gls{cpu} \& \gls{gpu} & Mid-complexity \\
\gls{mlp} & \gls{nn} & \gls{gpu} & High-complexity \\
\bottomrule
\end{tabular}
\end{table}


Random Forest is the only model excluded from one dataset combination: due to prohibitive training runtimes at 11 million instances, it is not evaluated on HIGGS in the main benchmark. It is, however, included in the dataset scaling experiment for subset sizes up to 500{,}000 instances, where runtimes remain manageable. This provides data on the point at which Random Forest becomes computationally impractical relative to the other models. All other model--dataset combinations are included in the main benchmark.

Together, the selected models cover a broad spectrum of algorithmic complexity, from a linear baseline to gradient boosting and deep learning, enabling a meaningful comparison of both predictive performance and energy efficiency across fundamentally different model types. The following section defines the metrics used to evaluate these dimensions.

\section{Evaluation Metrics}
\label{design_metrics}

To answer the research questions of this study, two categories of metrics were employed: classification performance and resource consumption.
\acrlong{acc} and \acrlong{wf1} were selected to measure predictive performance, as they provide a standardized basis for comparing models across datasets of varying class distributions. 
Energy consumption in \gls{kwh} and CO\textsubscript{2} emissions in \gls{kgco2eq} were chosen as the primary resource metrics, directly reflecting the ecological cost of model training.

Training time in seconds was included as a secondary resource metric, as it captures computational demand independently of hardware-specific power draw. 
Furthermore, inference time was measured across all experimental setups to account for the operational footprint. In large-scale deployment scenarios, the cumulative energy demand of repeated predictions can often rival or even exceed the initial training costs \cite{desislavovTrendsAIInference2023}.
Together, these metrics enable a direct assessment of whether gains in predictive performance justify the associated increase in resource consumption.

Inspired by the efficiency framing proposed in \citeauthor{schwartzGreenAI2020}, a custom composite metric was developed for this study, named the \gls{ewf1}. \Cref{eq:ewf1} defines this metric, where the numerator is \acrlong{wf1}:
\begin{equation}
\text{\gls{ewf1}} = \frac{\gls{wf1}}{1 + \lambda \cdot \ln\!\left(1 + \dfrac{c}{C_0}\right)}
\label{eq:ewf1}
\end{equation}
where $c$ is the measured CO\textsubscript{2} emissions in \gls{kgco2eq}, $C_0 = 10^{-3}$\,kg is a fixed reference emission value, and $\lambda$ controls the relative weight of ecological cost. The logarithmic penalty term $\ln(1 + c/C_0)$ grows monotonically with emissions but compresses extreme values, which is appropriate given that emissions span several orders of magnitude across model-dataset combinations.

The value $C_0 = 10^{-3}$\,kg was chosen because it lies close to the geometric mean of the observed emission range in this study, which extends from approximately $1.4 \times 10^{-5}$\,kg (Logistic Regression on Wine) to $1.5 \times 10^{-1}$\,kg (\gls{mlp} on HIGGS). Anchoring $C_0$ near this midpoint keeps the penalty term in an interpretable range of roughly $[0, 5]$ across all experiments. If $C_0$ were set much lower, virtually all emissions would produce a near-saturated penalty and the denominator would cease to discriminate; if it were set much higher, the penalty would approach zero for all but the most expensive runs and \gls{ewf1} would collapse towards \gls{wf1}. The reference is a fixed constant rather than a dataset-dependent quantity, making the metric portable: a model achieving a given \gls{wf1} at a given emission level receives the same \gls{ewf1} score regardless of which other models appear in the comparison pool.

The default weight $\lambda = 1$ assigns unit importance to the logarithmic ecological penalty relative to the \gls{wf1} numerator. This is the neutral starting point: ecological cost enters the denominator with the same order of magnitude as the performance term, without artificially amplifying or suppressing it. To verify that the conclusions of this study do not depend on this choice, \gls{ewf1} rankings were computed for $\lambda \in \{0.5,\, 1,\, 2\}$ across all three datasets. The resulting model orderings were identical in every case, confirming that the qualitative ranking is robust to the specific value of $\lambda$ and is not an artefact of the chosen weight. Higher scores indicate a more favorable trade-off between predictive performance and ecological cost.

It should be noted that $c$ is inherently location-dependent, as CO\textsubscript{2} emissions are tied to the carbon intensity of the local electricity grid. Consequently, \gls{ewf1} scores reflect the ecological efficiency of the specific experimental setup used in this study and may not generalize directly to other energy infrastructures.

Emissions are tracked exclusively over the \gls{cv} loop. Data loading and preprocessing are deliberately excluded from the measurement interval, as their cost scales with dataset size rather than model architecture and would therefore introduce noise into cross-model comparisons. This ensures that all reported emissions reflect model training and evaluation cost alone.

\section{Validation Strategy}
\label{design_validation}

To obtain reliable and generalizable performance estimates, a \gls{cv} strategy was employed rather than a single train-test split.

Specifically, K-Fold \gls{cv} was applied with $k=5$ folds, a value widely recommended in the literature as a practical balance between computational cost and estimation stability \cite{hastieElementsStatisticalLearning2017}. Each model is trained and evaluated five times on non-overlapping subsets of the data, and the final performance score is averaged across all folds. This approach is consistent across all models and datasets, ensuring comparability of results.

To ensure reproducibility, a fixed random state was used for all \gls{cv} splits, model initialization, and data shuffling. The specific value and its technical implementation are described in \Cref{implementation_sw_env}.

Because two of the three datasets exhibit class imbalance, stratified \gls{cv} is applied to ensure each fold preserves the original class distribution. This is directly coherent with the choice of \gls{wf1} as the primary metric, as described in \Cref{design_metrics}.

No separate held-out test set is used. The conventional alternative would be to reserve a fixed partition of the data for final evaluation and restrict \gls{cv} to the tuning phase. For Wine, with 178 instances, an 80/20 split would leave approximately 35 samples for testing, which is too small to produce a stable performance estimate. Five-fold \gls{cv} over all available instances evaluates each data point exactly once and therefore yields a lower-variance estimate than any fixed held-out subset of comparable size. For Credit and HIGGS, where each fold alone contains thousands to millions of instances, the variance of the performance estimate is negligible regardless of whether a separate test partition is used, and the choice has no practical consequence for the reported figures.

The remaining concern is that tuning and evaluation share the same folds, which introduces a slight optimistic bias through indirect data leakage: the selected hyperparameters are chosen to perform well on the validation folds, and those same folds contribute to the final reported score. This bias is, however, uniform across all models and datasets, because every configuration is subject to the identical procedure. When all competing models are subject to the same inflation, relative rankings are unaffected. The primary claims of this study are comparative rather than absolute, and the reported figures should be interpreted accordingly.

The theoretically exact remedy for this bias is nested \gls{cv}, in which an inner loop handles tuning and a fully independent outer loop estimates generalization error. On HIGGS, a five-fold outer loop would multiply \gls{hpo} evaluations fivefold per model--dataset combination, adding weeks of additional compute at substantial energy cost. Nested \gls{cv} is therefore rejected on ecological grounds consistent with the framing of this study, as discussed in \Cref{sec:theoretical_kfold}.

\section{Hyperparameter Tuning}
\label{design_hyperparameter}
Hyperparameter tuning was applied to all models except Logistic Regression. Without tuning, default parameters tend to favor certain model families over others, potentially skewing both performance and resource consumption comparisons \cite{bagnallUseDefaultParameter2017}. Logistic Regression is deliberately excluded: its role as a lightweight linear baseline depends on its simplicity, and tuning it would undermine that function.

Optuna was selected as the optimization framework, using \gls{tpe} as its search strategy. As discussed in \Cref{sec:theoretical_hpo}, \gls{bo}-based methods require substantially fewer evaluations than grid or random search to reach competitive configurations, which directly reduces the energy cost of the tuning process itself. Each run was configured to execute 40 trials per model--dataset combination, a budget chosen to balance search thoroughness against ecological overhead.

Tuning was performed separately for each dataset, as the datasets differ substantially in size, feature space, and class distribution. The optimization objective is \gls{wf1} in all cases, consistent with the primary evaluation metric defined in \Cref{design_metrics}. The concrete search spaces and implementation details are described in \Cref{implementation_tuning}.

A limitation of the fixed 40-trial budget is that it does not account for differences in search space dimensionality across models. Random Forest exposes five hyperparameters and XGBoost six, both in well-defined, low-dimensional spaces that \gls{tpe} can navigate reliably within 40 evaluations. The \gls{mlp} search space is larger and structurally more complex: it includes a discrete depth parameter (\texttt{n\_layers}, ranging from 1 to 4) that conditionally activates up to four additional layer-width dimensions, plus continuous parameters for dropout rate, learning rate, and batch size, yielding an effective dimensionality of five to eight depending on the sampled depth. Conditional search spaces of this kind require more evaluations for \gls{tpe} to build an accurate surrogate model. In addition, each \gls{mlp} trial on HIGGS involves training a neural network and is substantially more expensive than a comparable tree-based trial, so the 40-trial budget represents a greater practical constraint for the \gls{mlp} than for Random Forest or XGBoost. As a result, the \gls{mlp} may not have reached the same degree of optimization as the tree-based models, and the reported \gls{mlp} performance figures should be interpreted as a lower bound on what a more exhaustive search might achieve. Increasing the trial budget selectively for the \gls{mlp} would, however, break the comparability constraint: allocating more search effort to one model than to others would confound any observed performance difference with tuning thoroughness rather than isolating it as a property of the algorithm itself. The uniform budget is therefore a deliberate design choice to keep the search effort symmetric across all models.

\section{Experimental Protocol}
\label{sec:experimental_protocol}

The research questions are operationalized through five experiments. Each isolates a specific dimension of the energy-performance trade-off while holding the remaining factors constant. Implementation details for all five are described in \Cref{sec:experiments}.

The \textbf{main cross-dataset benchmark} trains all five models on all three datasets, yielding 14 model--dataset combinations (Random Forest on HIGGS is excluded, as described in \Cref{design_models}). It forms the empirical backbone of the study and addresses the overall energy-performance trade-off across model complexity and dataset scale.

The \textbf{dataset scaling experiment} holds everything constant except the number of training instances. All five models are included at subset sizes up to 500{,}000 instances; beyond that threshold Random Forest is excluded due to prohibitive runtimes, leaving four models for the larger subsets up to 11{,}000{,}000. Because all other variables are fixed, observed differences in performance and emissions can be attributed exclusively to dataset size. The gap between the 30{,}000-instance Credit dataset and the full 11-million-instance HIGGS dataset is bridged by the intermediate subset sizes in this experiment, which cover the range that is typical of many real-world tabular classification tasks. The experiment directly addresses RQ4 and, through the XGBoost \gls{cpu}/\gls{gpu} pairing, also feeds into RQ3.

The \textbf{\gls{mlp} architecture variation} isolates the effect of network depth and width by training a grid of 12 architectures on HIGGS: three widths (128, 256, 512 neurons) crossed with four depths (1--4 hidden layers). Non-architectural hyperparameters are fixed to neutral defaults to prevent any single configuration from being inadvertently favoured by the tuning results. This directly addresses RQ2.

The \textbf{temporal carbon intensity analysis} is purely retrospective: it re-weights the energy already recorded in the main benchmark against eight carbon intensity scenarios derived from one year of hourly \gls{api} data for the German grid. This quantifies the sensitivity of reported emissions to training time and location, and bounds the uncertainty introduced by CodeCarbon's static intensity factor.

The \textbf{lifecycle break-even analysis} combines the one-time training footprint from the main benchmark with per-prediction inference latency to estimate at what prediction volume cumulative inference emissions surpass training emissions. It addresses RQ5 without requiring any additional training runs.


